### 2025-04-09 chapters/5-event_shapes.tex
T-1 day to initial submission, nearly complete draft up to citations and small comments.

\chapter[Seeing through the Smog:\\\phantom{Chapter 5:|}Energy Flow and Weighted Correlations]{Seeing through the Smog\hfill\\Energy Flow and Weighted Correlators}
\markboth{\small \textsc{Chapter \thechapter}: \bf Seeing through the Smog: Energy Flow and Weighted Correlations}{}

@@

\iffalse
% TODO: move exposition

The flow of energy within hadronic jets is an indispensable probe of Quantum Chromodynamics (QCD)~\cite{PhysRevD.1.1416, PhysRevLett.39.1237, PhysRevLett.39.1587, PARISI197865, PhysRevD.20.2759, RAKOW198163}.
%

The discovery of hadronic jets in high-energy particle collisions is one of the most important milestones in human understanding of the microscopic universe, and has solidified quantum chromodynamics (QCD) as our fundamental theory of the strong nuclear force.
%
Jets were originally predicted \cite{Feynman:1969wa,Bjorken:1969wi,Drell:1969wb,Cabibbo:1970mh,Berman:1971xz}, and later measured \cite{Hanson:1975fe,Wiik:1979cq,Barber:1979yr,TASSO:1979zyf,PLUTO:1979dxn,JADE:1979rke,Ali:1979em,Hanson:1981em,Ali:2010tw}, as collimated sprays of hadronic radiation produced in high-energy collisions involving the strong interaction.
%
The quantification of the internal structure of jets, enabled by the excellent performance of the current generation of experimental detectors, has been a particularly fruitful area of study.
%
The study of jet substructure has led to the development of a large class of observables that characterize hadronic radiation in particle collisions, facilitating our understanding of the physical properties of the Standard Model of particle physics (SM) and beyond.

\fi

\begin{itemize}
    \item
We've discovered the good, the bad, and the ugly of jet grooming, especially tree-based jet grooming.

    \item
Instead of strictly removing soft particles, what about slightly ignoring them?
%
This leads us to energy-weighted correlators, which treat \textit{high-energy} objects as the most important.
%
After all, the high-energy structures within a jet are precisely those which, by asymptotic freedom, encode the physics of the partons of QCD.

    \item
Applications to \(\alpha_s\), \(m_t\), and beyond (QGP).
%
But for our purposes, we simply exposit how we may efficiently think about energy-weighted correlations, define them  in a way that avoids the infinities and seeming inconsistencies of quantum field theory, and impressionistically expound their usefulness in some very simple contexts without the burden of full realism.
\end{itemize}

@@

We have now seen that grooming provides a useful, but generically discontinuous, toolkit for extracting high-energy partonic information from jets at colliders.
%
The tree-based jet grooming algorithms that we can compare to the splitting structure of the partonic cascade have (albeit suppressed) discontinuities.

% TODO: change discussion
\sam{EWOCs except the EEC are discontinuous as long as the infra-red regulator is discontinuous...}
%
\sam{Discontinuity of infra-red regulators?}

In this chapter, we continue our search for soft-insensitive, continuous characterizations of partonic physics and jet substructure.
%
Rather than directly and completely removing radiation from the jet using a tree-based algorithm, however, we instead decide to \textit{partially ignore} soft radiation.
%
Operationally, this will mean multiplying the contributions of different emissions to the associated jet substructure observables by a factor proportional to their energy (or, more generally, a quantity that is largest for energetic particles and is zero for particles of zero energy).

@@

Importantly, we have left the sum on \(i,j\) a little bit vague.
%
What are we summing over?
%
The answer is that if we \(i\) and \(j\) to index particles, so that the \gls{ewoc} captures pair-wise particle correlations, we quickly run into trouble.
%
In particular, collinear safety requires that particle-level \glspl{ewoc} can \textit{only} encode angular correlations within a jet:
%
\(\mathcal{O}_{ij} = \theta_{ij}\) or \(\mathcal{O}_{ij} = \Delta R_{ij}\).
%
In order to obtain \glslink{irc-safety}{IRC safe} observable, we must either study only \textit{angular correlations} or \textit{give up on particle-level correlations} in favor of correlations between collective degrees of freedom.%
\footnote{
    We will also see that the requirement of \gls{continuity} is again quite difficult to achieve, and that only energy correlators -- particle-level \glspl{ewoc} involving angular correlations -- are continuous.
    %
    Nonetheless, the study of \glspl{ewoc} will lead to results which are \textit{more} robust to hadronization.
}

@@

% -----------------------------------
% Picturebook figure
% -----------------------------------
\reusefigure[ht]{picturebook_eventshapes}

@@

Among jet substructure observables, the Energy-Energy Correlator (EEC) \cite{Basham:1978bw,Basham:1978zq,Basham:1979gh} is particularly noteworthy:
%
it is conceptually simple, insensitive to low-energy radiation, and naturally separates physics at different scales.
%
Physically, the EEC measures the mean correlation in the energy passing through any two detectors separated by an angle \(\Delta\), averaged over many particle collisions:
%
\(
    \text{EEC}(\Delta)
    \propto
    \left\langle
        E_\text{det.\.1} \,\,\, E_\text{det.\.2}
    \right\rangle
\).%
\footnote{
    More concretely, this means that the contribution of each particle in an event to the EEC is weighted by the particle's energy,
    %
    see \Eq{eec_defn}.
    %
    This energy weighting mitigates the effects of the low-energy physics of QCD that complicates the theoretical description of other jet substructure observables.
}
%
From the birth of the EEC to its use in recent collider studies, the simplicity of the EEC has enabled highly accurate perturbative computations \cite{Clay:1995sd,Glover:1994vz,Kramer:1996qr,DelDuca:2016csb,Gituliar:2017umx,Dixon:2018tpg,Dixon:2018qgp,Henn:2019gkr,Luo:2019nig,Gao:2020vyx,Neill:2022lqx,Lee:2023npz,Kramer:1995qh,deFlorian:2004mp,Banfi:2002vw,Tulipant:2017ybb,Moult:2018jzp,Korchemsky:2019nzm,Dixon:2019uzg,Gao:2019ojf,Luo:2019hmp,Luo:2019bmw,Moult:2019vou,Li:2020bub,Ebert:2020sfi,Li:2021txc,Duhr:2022yyp,Chen:2023zlx,Gao:2023ivm} and has facilitated measurements of the strong coupling constant \(\alpha_s\) \cite{Martin:1986uq,DELPHI:1990sof,SLD:1994yoe,ATLAS:2015yaa,ATLAS:2017qir,dEnterria:2018cye,Kardos:2018kqj,Ali:2020ksn,dEnterria:2022hzv,ATLAS:2023tgo} including the most precise jet substructure determinations of \(\alpha_s\) to date \cite{CMS:2024mlf}.
%
The EEC has been investigated as a probe of jets traveling through the quark-gluon plasma (QGP) produced in heavy-ion collisions \cite{Lokhtin:2004tx,Lokhtin:2006dp,Andres:2022ovj,Barata:2023zqg,Andres:2023xwr,Yang:2023dwc,Barata:2023vnl,Barata:2023bhh,Barata:2024nqo} and of correlations in nuclear physics \cite{Karapetyan:2019fst,Liu:2022wop,Liu:2023aqb,Kang:2023gvg,Cao:2023oef}.
%
The study of the EEC has also shed light on the structure of non-perturbative QCD \cite{Nason:1995np,Korchemsky:1997sy,Korchemsky:1999kt,Dokshitzer:1999sh,Chen:2020vvp,Jaarsma:2023ell,Schindler:2023cww,Lee:2024esz}
and of quantum field theory in general \cite{Richards:1983sr,Sveshnikov:1995vi,Hofman:2008ar,Hatta:2012kn,Belitsky:2013bja,Belitsky:2013ofa,Belitsky:2013xxa,Belitsky:2014zha,Korchemsky:2015ssa,Goncalves:2014ffa,Farnsworth:2015hum,Hofman:2016awc,Kravchuk:2018htv,Kologlu:2019mfz,Chang:2020qpj,Korchemsky:2021htm,Caron-Huot:2022eqs,Chen:2023wah,Chicherin:2023gxt,Chicherin:2024ifn}.
%
Finally, while most jet substructure observables are dominated by contributions from two-particle correlations, the EEC can be naturally extended to multi-point energy correlators which can characterize correlations between an arbitrary number of particles.
%
Recent theoretical work suggests that multi-point generalizations of the EEC may be used to extract the mass of the top quark at the LHC \cite{Procura:2022fid,Holguin:2022epo,Holguin:2023bjf,Pathak:2023tmy,Xiao:2024rol,Holguin:2024tkz}.

\sam{Cite \cite{Ricci:2022htc} somewhere}

@@

\begin{itemize}
    \item
        % TODO: EEC is continuous
    Argue that, on a jet-by-jet level, each jet has a continuous contribution (as a map from energy flows to 1D distributions) though we will be working at the level of average correlations/distributions, as before
    %
    (a small change in the jet leads to a small change in the EEC by the Wasserstein metric/spectral distance -- cite Jesse and Andrew)
\end{itemize}

@@

\remark{}{
    Notably, the sum over pairs of particles \(i\) and \(j\) includes terms with \(i = j\), which contribute only when \(\chi = 0\):
    \begin{align}
        \dv{\Sigma}{\cos\chi}{}\le(\chi\ri)
        &=
        \notag
        2 \int \dd\sigma
        \sum_{i < j} \frac{E_i}{Q} \frac{E_j}{Q}
        \delta\le(\chi \, - \, \frac{1-\cos(\theta_{ij})}{2}\ri)
        +
        \int \dd\sigma
        \sum_{i} \frac{E_i}{Q} \frac{E_i}{Q}
        \delta\le(\chi\ri)
        \,,
    \end{align}
    where we have used \(\sum_{i, j} = \sum_i + 2 \sum_{i<j}\).
    %
    The terms in the second sum, \(\sum_i\), are referred to as \vocab{contact terms}.
    also known as \vocab{contact} terms
    %
    The contact terms ensure that the EEC integrates to 1:
    \begin{align}
        \label{eq:app:EEC:normalization}
        \int_0^{\chi_\text{max}}
        \frac{\dd \Sigma_\text{EEC}}{\dd \chi}
        =
        \int \frac{\dd\sigma}{\sigma}
        \,
        \sum_{\text{particles } i,\,j}
        \frac{E_i E_j}{E_\text{tot}^2}
        =
        \int \frac{\dd\sigma}{\sigma}
        \,
        \frac{\le(\sum_{\text{particles } i} E_i\ri)^2}{E_\text{tot}^2}
        =
        1
        \,.
    \end{align}
    Similar arguments hold for the energy correlators discussed in \Sec{new-angles} as well, where analogous contact terms ensure that higher-point energy correlators also integrate to one.
}

@@

\remark{eec-energyflow}{
    The EEC of \Eq{eec-real-angle} in a theory with massless particles (or in the high-energy limit \(m^2 / Q^2 \ll 1\) for all particle masses \(m\)) can also be expressed in terms of the energy flow operators of \Def{energy-flow} as
    \begin{align}
        \label{eq:eec-energy-flow}
        \frac{\dd \Sigma_\text{eec}}{\dd\chi}
        =
        \int
        \,
        \dd\Omega_1 \dd\Omega_2
        \,
        \le\langle
            \frac{1}{Q^2}
            \mathcal{E}(\hat{n}_1)
            \mathcal{E}(\hat{n}_2)
        \ri\rangle
        \,\,
        \delta\le(
            \chi - \frac{1 - \hat{n}_1\cdot\hat{n}_2}{2}
        \ri)
        \,,
    \end{align}
    where the expectation value indicates a quantum mechanical average, potentially over an ensemble of initial states, as discussed in Remark~\ref{rem:eec-average}.
    %
    \sam{not done, but want something like \(\langle A \rangle = \sum_{\substack{\text{ensemble of}\\\text{final states}}} \bra{f}A\ket{f} = \Tr \mathbb{P} S \rho S^\dagger \mathbb{P}^\dagger A\), where \(\mathbb{P}\) is a projection onto the states defined by \(\dd\sigma\).}
    %
    For the definition of \Eq{eec-deltaR}, measuring correlations of distance \(\Delta R\) in the rapidity-azimuth plane, the delta function of \Eq{eec-energy-flow} must be replaced by
    \begin{align}
        \delta\le(R_{ij} - \Delta R\le(\hat{n}_1, \hat{n}_2\ri)\ri)
        \,.
    \end{align}
    In practice, one will also add cuts on the radiation considered in the EEC or consider the EEC only within a specific jet, which requires the use of additional constraints (for example, the use of Heaviside Theta functions in the integrand, dependent on the energy flow itself, which enforce that \(\hat{n}_1\) and \(\hat{n}_2\) lie within the same jet).
    %
    While these additional constraints may be complicated functionals of the energy flow operator, they may generally be disregarded in the collinear limit that we consider.
    % TODO: refine/incorporate
    %Energy correlator observables~\cite{Sterman:1975xv, Basham:1977iq, Basham:1978bw, Basham:1978zq, Basham:1979gh} are particularly powerful tools for understanding energy flow both theoretically and experimentally~\cite{Mazzilli:2024ots,CMS:2024mlf,Tamis:2023guc}.
    %%
    %Since energy correlators can be described directly in terms of field-theoretic energy flow operators~\cite{Sveshnikov:1995vi, Tkachov:1995kk, Korchemsky:1999kt, Lee:2006nr, Hofman:2008ar, Belitsky:2013xxa, Belitsky:2013bja, Kravchuk:2018htv}, one can use sophisticated theoretical techniques, including the powerful technology of conformal field theories~\cite{Hofman:2008ar, Kologlu:2019mfz}, to extract rich information about jet substructure, especially in the collinear limit~\cite{Dixon:2019uzg, Chen:2020vvp, Chen:2019bpb, Chen:2020adz, Chen:2021gdk, Schindler:2023cww, Gao:2023ivm, Chen:2023zlx, Chicherin:2024ifn, Chen:2024nyc, Chang:2022ryc,Budhraja:2024xiq,Budhraja:2024tev,Lee:2024jnt}.
}

@@

    The factorization of the EEC in the collinear limit is discussed in \Reff{}.
    %
    Physically, EEC factorization occurs neither the hard function nor the jet function contain any information about correlations between particles of different jets, so we can only use them to probe angular correlations of pairs of particles within a single jet.
    %
    However, inter-jet correlations become invisible in the collinear limit, and \Eq{factorization_defn} becomes a powerful approximation.

@@

\begin{exercise}
    Argue that the hard function for the process \(e^+ e^- \to\,\)hadrons is simply
    \begin{align}
        H_i(z)
        =
        \delta\le(z - \frac{1}{2}\ri)
        \le(\delta_{i\,q} + \delta_{i\,\bar{q}}\ri)
        \,,
    \end{align}
    where \(z\) indicates an energy fraction relative to the center-of-mass energy of the entire event, at \(\mathcal{O}(\alpha_s^0)\).
    %
    Conclude that to obtain leading results for the EEC in the collinear limit, we need only the EEC quark jet function.
\end{exercise}

\remark{}{
    As discussed in , The EEC is infra-red/collinear (IRC) safe (as we discuss further in \Sec{ewoc-irc} in the context of general \glspl{ewoc}) and can therefore be computed in perturbation theory.
    %
    In particular, the use of energy weighting ensures that the EEC is unchanged under the addition of zero-energy radiation (infra-red safety), and the energy weighting together with the dependence of the delta function solely on the angles of particles ensures that it is unchanged under exactly collinear splittings of final-state particles (collinear safety).
    %
    Furthermore, the energy weighting in the integrand also suppresses non-perturbative corrections arising from soft physics to perturbative calculations of the EEC.
    %
    The IRC robustness of the EEC is one it is useful tool for the extraction of the strong coupling constant from particle collision data.
}

@@

First, we recall the important universal pieces of the collinear limit that will be important to our discussion \cite{Altarelli:1977zs,PhysRevD.9.980,PhysRevD.46.1980,Ritzmann:2014mka}.
%
In dimensional regularization with \(d = 4 - 2\epsilon\) spacetime dimensions, the differential cross-section for the collinear limit of the amplitude for \(q^* \to q\,g\) is given by
\begin{align}
    \label{eq:collinear_xsec}
    \frac{\dd \sigma_{qg}}{\dd \Phi_{qg}}
    =
    \mu^{2\epsilon}\,
    e^{\epsilon \gamma_E}
    \,
    \frac{2^{3-2\epsilon} \pi^{1-\epsilon}  \alpha_s}{s} P_{qg\leftarrow q}(z)
    ,
\end{align}
where \(s\) is the invariant mass of the final-state quark-gluon pair (the virtuality of the intermediate quark), \(z = E_q/Q\) is the energy fraction of the outgoing quark, and we use the Lorentz-invariant phase space measure for the outgoing quark and gluon in the collinear limit:
\begin{align}
    \label{eq:collinear_phase_space}
    \dd \Phi_{qg}
    =
    \frac{
        {(4\pi)}^{\epsilon-2}
        s^{-\epsilon}
        {(1-z)}^{-\epsilon} z^{-\epsilon}
    }
    {\Gamma(1-\epsilon)}
    \,
    \dd z \, \dd s
    .
\end{align}
%
\sam{reference collinear limit section}

\begin{proposition}{}{}
 The differential EEC for any process containing an intermediate quark (or anti-quark) should contains a universal piece of the form
\begin{align}
    \label{eq:universal_differential_eec}
    \frac{\dd \Sigma_\text{EEC}}{\dd \chi}
    (\chi \to 0)
    \supset
    \le[
    \frac{3\ascf}{4\chi}
    \ri]_+
    \,,
\end{align}
up to a numerical, process dependent coefficient encoding the average number density and energy fraction of quarks (or anti-quarks).
\end{proposition}

>>>>

\chapter[Seeing through the Smog:\\\phantom{Chapter 5:|}Energy Flow and Weighted Correlations]{Seeing through the Smog:\hfill\\Energy Flow and Weighted Correlators}
\markboth{\small \textsc{Chapter \thechapter}: \bf Seeing through the Smog: Energy Flow and Weighted Correlations}{}

@@

%We've now discovered the good, the bad, and the ugly of jet grooming;
%%
%we've overcome the soft discontinuities of hard-cutoff jet grooming, and uncovered that all tree-based jet grooming algorithms suffer from \glslink{angular-discontinuity}{angular} \glspl{clustering-discontinuity}.
%%
%In this chapter, we continue our search for continuous characterizations of jet substructure by finding soft-continuous methods for \textit{gently ignoring} soft particles -- rather than strictly removing them -- with the framework of \vocab{\gls{energy-weighted-correlations}}.

In this chapter, we continue characterizing the substructure of jets, this time with energy-weighted correlations.
%
Much like jet grooming, \gls{energy-weighted-correlations} uncover high energy structures within jets;
%
however, they do so completely removing the influence of low energy radiation, leading to formal and phenomenological benefits that we will see throughout this chapter.

In \Sec{eec}, we exposit the very basics of the most common energy-weighted correlation, the \glslink{eec}{energy-energy correlator (EEC)} -- a probe of angular correlations between high energy structures within jets, and gain an understanding of how the EEC characterizes the internal viscera of jets using the technology of \gls{parton-to-parton} discussed in \Sec{p2p-fragmentation}.
%
In \Sec{new-angles}, we explore higher-particle generalizations of the EEC, and the work of myself and my collaborators in creating visually intuitive and computationally efficient parametrizations which are more suitable for use on realistic datasets than previous parametrizations in the literature.
%
Finally, in \Sec{ewocs}, we explore the more general framework of \glslink{ewoc}{energy-weighted observable correlations (EWOCs)}, introduced by myself and Wouter Waalejwin, for probing high-energy structures within jets by expanding the philosophy of the EEC to non-angular correlations.
%
We will see that the mass EWOC, which captures pair-wise mass correlations between subjets within a jet, is a powerful tool for extracting mass scales associated with jet production which has phenomenological benefits even over jet grooming and the EEC.

@@

We have now seen that grooming provides a useful, but generically discontinuous, toolkit for extracting high-energy partonic information from jets at colliders.
%
In this chapter, we continue our search for soft-insensitive, continuous characterizations of partonic physics and jet substructure.
%
Rather than directly and completely removing radiation from the jet using a tree-based algorithm, however, we instead decide to \textit{partially ignore} soft radiation.
%
Operationally, this will mean multiplying the contributions of different emissions to the associated jet substructure observables by a factor proportional to their energy (or, more generally, a quantity that is largest for energetic particles and is zero for particles of zero energy).

@@

Importantly, we have left the sum on \(i,j\) a little bit vague.
%
What are we summing over?
%
The answer is that if we \(i\) and \(j\) to index particles, so that the \gls{ewoc} captures pair-wise particle correlations, we quickly run into trouble.
%
In particular, collinear safety requires that particle-level \glspl{ewoc} can \textit{only} encode angular correlations within a jet:
%
\(\mathcal{O}_{ij} = \theta_{ij}\) or \(\mathcal{O}_{ij} = \Delta R_{ij}\).
%
In order to obtain \glslink{irc-safety}{IRC safe} observable, we must either study only \textit{angular correlations} or \textit{give up on particle-level correlations} in favor of correlations between collective degrees of freedom.%
\footnote{
    We use subjets as proxies for collective degrees of freedom so that our energy-weighted correlations are not fully continuous \glslink{continuity}{continuous} (the number of subjets in a jet, and the discrete yes/no choice of whether a particle belongs to a given subjet, are both discontinuous).
    %
    The only fully continuous substructure observables of this chapter are the EEC and its multi-particle generalizations.
    %
    Nonetheless, the study of \glspl{ewoc} will lead to results which are \textit{more} robust to hadronization.
}

@@

% -----------------------------------
% Picturebook figure
% -----------------------------------
\reusefigure[t]{picturebook_eventshapes}

@@

Among jet substructure observables, the Energy-Energy Correlator (EEC) \cite{Basham:1978bw,Basham:1978zq,Basham:1979gh} is particularly noteworthy:
%
it is conceptually simple, insensitive to low-energy radiation, and naturally separates physics at different scales.
%
Physically, the EEC measures the mean correlation in the energy passing through any two detectors separated by an angle \(\Delta\), averaged over many particle collisions:
%
\(
    \text{EEC}(\Delta)
    \propto
    \left\langle
        E_\text{det.\.1} \,\,\, E_\text{det.\.2}
    \right\rangle
\).%
\footnote{
    More concretely, this means that the contribution of each particle in an event to the EEC is weighted by the particle's energy,
    %
    see \Eq{eec_defn}.
    %
    This energy weighting mitigates the effects of the low-energy physics of QCD that complicates the theoretical description of other jet substructure observables.
}
%
From the birth of the EEC to its use in recent collider studies, the simplicity of the EEC has enabled highly accurate perturbative computations \cite{Clay:1995sd,Glover:1994vz,Kramer:1996qr,DelDuca:2016csb,Gituliar:2017umx,Dixon:2018tpg,Dixon:2018qgp,Henn:2019gkr,Luo:2019nig,Gao:2020vyx,Neill:2022lqx,Lee:2023npz,Kramer:1995qh,deFlorian:2004mp,Banfi:2002vw,Tulipant:2017ybb,Moult:2018jzp,Korchemsky:2019nzm,Dixon:2019uzg,Gao:2019ojf,Luo:2019hmp,Luo:2019bmw,Moult:2019vou,Li:2020bub,Ebert:2020sfi,Li:2021txc,Duhr:2022yyp,Chen:2023zlx,Gao:2023ivm} and has facilitated measurements of the strong coupling constant \(\alpha_s\) \cite{Martin:1986uq,DELPHI:1990sof,SLD:1994yoe,ATLAS:2015yaa,ATLAS:2017qir,dEnterria:2018cye,Kardos:2018kqj,Ali:2020ksn,dEnterria:2022hzv,ATLAS:2023tgo} including the most precise jet substructure determinations of \(\alpha_s\) to date \cite{CMS:2024mlf}.
%
The EEC has been investigated as a probe of jets traveling through the quark-gluon plasma (QGP) produced in heavy-ion collisions \cite{Lokhtin:2004tx,Lokhtin:2006dp,Andres:2022ovj,Barata:2023zqg,Andres:2023xwr,Yang:2023dwc,Barata:2023vnl,Barata:2023bhh,Barata:2024nqo} and of correlations in nuclear physics \cite{Karapetyan:2019fst,Liu:2022wop,Liu:2023aqb,Kang:2023gvg,Cao:2023oef}.
%
The study of the EEC has also shed light on the structure of non-perturbative QCD \cite{Nason:1995np,Korchemsky:1997sy,Korchemsky:1999kt,Dokshitzer:1999sh,Chen:2020vvp,Jaarsma:2023ell,Schindler:2023cww,Lee:2024esz}
and of quantum field theory in general \cite{Richards:1983sr,Sveshnikov:1995vi,Hofman:2008ar,Hatta:2012kn,Belitsky:2013bja,Belitsky:2013ofa,Belitsky:2013xxa,Belitsky:2014zha,Korchemsky:2015ssa,Goncalves:2014ffa,Farnsworth:2015hum,Hofman:2016awc,Kravchuk:2018htv,Kologlu:2019mfz,Chang:2020qpj,Korchemsky:2021htm,Caron-Huot:2022eqs,Chen:2023wah,Chicherin:2023gxt,Chicherin:2024ifn}.
%
It also been put forward as a way to search for new physics in hadronic decays of electroweak bosons~\cite{Ricci:2022htc}.
%
Finally, while most jet substructure observables are dominated by contributions from two-particle correlations, the EEC can be naturally extended to multi-point energy correlators which can characterize correlations between an arbitrary number of particles.
%
Recent theoretical work suggests that multi-point generalizations of the EEC may be used to extract the mass of the top quark at the LHC \cite{Procura:2022fid,Holguin:2022epo,Holguin:2023bjf,Pathak:2023tmy,Xiao:2024rol,Holguin:2024tkz}.

@@

@@

\remark{}{
    Notably, the sum over pairs of particles \(i\) and \(j\) includes terms with \(i = j\), which contribute only when \(\chi = 0\):
    \begin{align}
        \dv{\Sigma}{\cos\chi}{}\le(\chi\ri)
        &=
        \notag
        2 \int \dd\sigma
        \sum_{i < j} \frac{E_i}{Q} \frac{E_j}{Q}
        \delta\le(\chi \, - \, \frac{1-\cos(\theta_{ij})}{2}\ri)
        +
        \int \dd\sigma
        \sum_{i} \frac{E_i}{Q} \frac{E_i}{Q}
        \delta\le(\chi\ri)
        \,,
    \end{align}
    where we have used \(\sum_{i, j} = \sum_i + 2 \sum_{i<j}\).
    %
    The terms in the second sum, \(\sum_i\), are referred to as \vocab{contact terms}.
    %
    The contact terms ensure that the EEC integrates to 1:
    \begin{align}
        \label{eq:app:EEC:normalization}
        \int_0^{\chi_\text{max}}
        \frac{\dd \Sigma_\text{EEC}}{\dd \chi}
        =
        \int \frac{\dd\sigma}{\sigma}
        \,
        \sum_{\text{particles } i,\,j}
        \frac{E_i E_j}{E_\text{tot}^2}
        =
        \int \frac{\dd\sigma}{\sigma}
        \,
        \frac{\le(\sum_{\text{particles } i} E_i\ri)^2}{E_\text{tot}^2}
        =
        1
        \,.
    \end{align}
    Similar arguments hold for the energy correlators discussed in \Sec{new-angles} as well, where analogous contact terms ensure that higher-point energy correlators also integrate to one.
}

@@

\remark{eec-energyflow}{
    The EEC of \Eq{eec-real-angle} in a theory with massless particles (or in the high-energy limit \(m^2 / Q^2 \ll 1\) for all particle masses \(m\)) can also be expressed in terms of the energy flow operators of \Def{energy-flow} as
    \begin{align}
        \label{eq:eec-energy-flow}
        \frac{\dd \Sigma_\text{eec}}{\dd\chi}
        =
        \int
        \,
        \dd\Omega_1 \dd\Omega_2
        \,
        \le\langle
            \frac{1}{Q^2}
            \mathcal{E}(\hat{n}_1)
            \mathcal{E}(\hat{n}_2)
        \ri\rangle
        \,\,
        \delta\le(
            \chi - \frac{1 - \hat{n}_1\cdot\hat{n}_2}{2}
        \ri)
        \,,
    \end{align}
    where the expectation value indicates a quantum mechanical average, potentially over an ensemble of initial states.
    %
    When measuring correlations of distance \(\Delta R\) in the rapidity-azimuth plane -- i.e. in proton-proton collisions -- the delta function of \Eq{eec-energy-flow} must be replaced by
    \begin{align}
        \delta\le(R_{ij} - \Delta R\le(\hat{n}_1, \hat{n}_2\ri)\ri)
        \,.
    \end{align}
    In practice, one will also add cuts on the radiation considered in the EEC or consider the EEC only within a specific jet, which requires the use of additional constraints (for example, the use of Heaviside Theta functions in the integrand, dependent on the energy flow itself, which enforce that \(\hat{n}_1\) and \(\hat{n}_2\) lie within the same jet).
    %
    While these additional constraints may be complicated functionals of the energy flow operator, they may generally be disregarded in the collinear limit that we consider.
}

@@

    The factorization of the EEC in the collinear limit is discussed in \Reff{Dixon:2019uzg}.
    %
    Physically, EEC factorization occurs neither the hard function nor the jet function contain any information about correlations between particles of different jets, so we can only use them to probe angular correlations of pairs of particles within a single jet.
    %
    However, inter-jet correlations become invisible in the collinear limit, and \Eq{factorization_defn} becomes a powerful approximation.

\begin{exercise}
    Use the formal definitions of the hard function and jet function above to derive \Lem{eec-factorization}.
    %
    Explain the factor of \(z^2\).
\end{exercise}

@@

\begin{exercise}
    \label{ex:ee-hard-function}
    Argue that the LO hard function for the process \(e^+ e^- \to\,\)hadrons is simply
    \begin{align}
        H_i(z)
        =
        \delta\le(z - \frac{1}{2}\ri)
        \le(\delta_{i\,q} + \delta_{i\,\bar{q}}\ri)
        \,,
    \end{align}
    where \(z\) indicates an energy fraction relative to the center-of-mass energy of the entire event, at \(\mathcal{O}(\alpha_s^0)\).
    %
    Conclude that to obtain leading results for the EEC in the collinear limit, we need only the quark jet function for the EEC.
\end{exercise}

\remark{}{
    As discussed in , The EEC is \glslink{irc-safety}{IRC safe} (as we discuss further in \Sec{ewoc-irc} in the context of general \glspl{ewoc}) and can therefore be computed in perturbation theory.
    %
    In particular, the use of energy weighting ensures that the EEC is unchanged under the addition of zero-energy radiation (infra-red safety), and the energy weighting together with the dependence of the delta function solely on the angles of particles ensures that it is unchanged under exactly collinear splittings of final-state particles (collinear safety).
    %
    Furthermore, the energy weighting in the integrand also suppresses non-perturbative corrections arising from soft physics to perturbative calculations of the EEC.
    %
    The IRC robustness of the EEC is one it is useful tool for the extraction of the strong coupling constant from particle collision data.
}

@@

First, we recall the important universal features of the collinear limit of \gls{qcd} that will be important to our discussion.
%
Borrowing from \Sec{universal-features} (and in turn from \cite{Altarelli:1977zs,PhysRevD.9.980,PhysRevD.46.1980,Ritzmann:2014mka}), the differential cross-section for the collinear limit of the amplitude for \(q^* \to q\,g\) in \(d = 4 - 2\epsilon\) spacetime dimensions is given by
\begin{align}
    \label{eq:collinear_xsec}
    \frac{\dd \sigma_{qg}}{\dd \Phi_{qg}}
    =
    \mu^{2\epsilon}\,
    e^{\epsilon \gamma_E}
    \,
    \frac{2^{3-2\epsilon} \pi^{1-\epsilon}  \alpha_s}{s} P_{qg\leftarrow q}(z)
    ,
\end{align}
where \(s\) is the invariant mass of the final-state quark-gluon pair (the virtuality of the intermediate quark), \(z = E_q/Q\) is the energy fraction of the outgoing quark, and we use the Lorentz-invariant phase space measure for the outgoing quark and gluon in the collinear limit:
\begin{align}
    \label{eq:collinear_phase_space}
    \dd \Phi_{qg}
    =
    \frac{
        {(4\pi)}^{\epsilon-2}
        s^{-\epsilon}
        {(1-z)}^{-\epsilon} z^{-\epsilon}
    }
    {\Gamma(1-\epsilon)}
    \,
    \dd z \, \dd s
    .
\end{align}

\begin{proposition}{The (differential) NLO EEC Jet Function}{nlo-eec-jet-function}
The differential EEC for any process containing a massless intermediate quark (or anti-quark) contains a universal piece of the form
\begin{align}
    \label{eq:universal_differential_eec}
    \frac{\dd \Sigma_\text{EEC}}{\dd \chi}
    (\chi \to 0)
    \supset
    \le[
    \frac{3\ascf}{\chi}
    \ri]_+
    \,,
\end{align}
up to a numerical, process dependent coefficient encoding the average number density and energy fraction of quarks (or anti-quarks).

\Eq{universal_differential_eec} is the EEC associated with a single splitting within a quark jet, normalized to the energy of the quark.
%
Therefore, it is the \(\mathcal{O}(\alpha_s)\) (NLO) jet function for the \textit{differential} EEC \cite{Dixon:2019uzg}.
\end{proposition}

\begin{proof}
    Let \(Q\) denote the energy of an outgoing, off-shell quark which undergoes a collinear splitting into a final-state quark and gluon.
    %
    We use use \(z = E_q / Q\) and \((1-z) = E_g/Q\) to denote the energy fractions of the final-state quark and gluon, respectively, and \(\chi_{qg} = (1 - \cos\theta_{qg})/2 \approx \theta_{qg}^2/4\) to describe their angular separation.
